Gravitational Lens Recovery with \Glass: Measuring the mass profile and shape of a lens

### 2014-01-29 limits/ms.tex
Fixed jpc affiliation.

\title[Lens Recovery with \Glass]{Gravitational Lens Recovery with \Glass: Measuring the mass profile and shape of a lens}
\author[J.~P.~Coles~et~al.]{%
Jonathan P.~Coles,$^{1,3}$\thanks{jonathan@exascale-computing.eu}%
\newauthor%
Justin I.~Read$^2$,%
\newauthor%
and Prasenjit Saha$^3$%
%
\\
$^1$Exascale Research Computing Lab, Universit\'e de Versailles Saint-Quentin-en-Yvelines,\\$\quad$55 Avenue des Etats Unis, 78000, Versailles, France\\
$^2$Department of Physics, University of Surrey, Guildford, Surrey, GU2 7XH, United Kingdom \\
$^3$Physik-Institut, Universit\"at Z\"urich, 190 Winterthurerstrasse, 8057, Z\"urich, Switzerland
}

@@

To see why lens modelling details are of crucial importance, let us
recall the essential quantities that appear in lensing (see also \S\ref{sec:theory} for a more detailed exposition). First we have the distances. Let $D_L$, $D_S$, $D_{LS}$ be the
angular-diameter distances to the lens, source, and from lens to
source; these are all proportional to $c/H_0$ but have factors
that depend on the particular choice of cosmology\footnote{Here, $c$ is the speed of light in vacuo and $H_0$ is the Hubble parameter.}.  Typically:
%
\begin{equation}
D_L \approx z_L \frac c{H_0}\ \hbox{and}\ \frac{D_S}{D_{LS}} \sim 1.
\end{equation}
where $z_L$ is the redshift of the lens. 
%
For multiple images, the sky-projected density must exceed the critical
density in some region:
%
\begin{equation}
\Sigma_\mathrm{crit} = \frac{c^2}{4\pi G D_L} \sim \frac{1\rm\,kg\,m^{-2}}{z_L}
\end{equation}
where $G$ is Newton's gravitational constant. 
%
The separation between the lensed images is of order
the Einstein radius $\theta_E$, which is related to
the mass by:
%
\begin{equation}
\theta_E \sim \frac{R_G}{D_L} \frac{D_{LS}}{D_S}
\end{equation}
%
where $R_G = GM/c^2$ (with $M$ the projected mass enclosed within $\theta_E$) is the gravitational radius. If the source is a
quasar or otherwise rapidly variable, a time delay $\Delta t$ in the variability
will be present where:
%
\begin{equation}
\Delta t \sim R_G/c \,.
\end{equation}
%
So in principle, one can not only measure the mass of the lens, one can use the
dependence on the cosmology-dependent $D$ factors to extract the cosmological
model and all its parameters.  \cite{1937ApJ....86..217Z} drew attention to the
former, and \cite{1964MNRAS.128..307R,1966MNRAS.132..101R} pointed out the
latter, all long before lenses were discovered.  The difficulty with actually
doing this, however, became apparent soon after the discovery of the first lens
by \cite{1979Natur.279..381W}.  In the first ever paper on lens modelling,
\cite{1981ApJ...244..736Y} found that many plausible mass distributions could
reproduce the data.

@@

\subsection{A post-processing module for stellar kinematics}\label{sec:glasskinematics} 
Similarly to the null-space constraint (\secref{sec:glassextraimages}), stellar
kinematic constraints constitute a non-linear prior on the mass map and must be
applied in post-processing. We sweep through the model ensemble performing an
Abel deprojection to determine $\Mddd(r)$ from the projected surface density
$\Sigma(R)$ assuming spherical symmetry
\citep[e.g.][]{2008gady.book.....B,2008MNRAS.390.1647B}: 
%
\begin{eqnarray} 
    \Mddd(r) & = & 2\pi \int_0^r R \Sigma(R) dR - 4r^2 \int_0^{\pi/2} \Sigma\left(x\right) \left(\frac{1}{\cth[2]} \right. \nonumber \\ 
    & & \left. - \frac{\sth}{\cth[3]} \arctan\left(\frac{\cth}{\sth}\right) \right) \dth
\end{eqnarray}
%
where $M(<r)$ is the projected enclosed mass evaluated at 3D radius $r$; and $x = r/\cth$.

@@

\begin{figure*}
\includegraphics[width=0.33\textwidth]{BCQuadR1a_TmS-kappa-a.pdf}
\includegraphics[width=0.33\textwidth]{BCQuadR1a_TmS-kappa-b.pdf}
\caption{ \textbf{Left:} The mock data distribution for \mockBC{} projected onto a coarse grid.
  \textbf{Right:} The recovered ensemble average $\kappa$ distribution for the single quad. The contours are
  logarithmic base 10 values, where level 0 corresponds to the critical density. Contours below the critical density
  are drawn with dashed lines.}
\label{2d mass reconstruction}
\end{figure*}

@@

We can also more directly probe the recovery of the shape of the mass
distribution by defining a global measure of lens shape:
%
\begin{equation} 
    s \equiv \lambda_1/\lambda_2
\end{equation} 
%
where $\lambda_1$ and $\lambda_2$ are the eigenvalues of the 2D inertia tensor:
%
% \begin{gather}
%    I_{xx} = \sum_{\vec\theta} M(\vec\theta) \theta^2_y \quad ; \quad
%    I_{yy} = \sum_{\vec\theta} M(\vec\theta) \theta^2_x \\
%    I_{xy} = I_{yx} = - \sum_{\vec\theta} M(\vec\theta) \theta_x \theta_y \nonumber
%    \label{eqn:inertia}
% \end{gather}
%
\begin{equation}
\left(
\begin{matrix}
   \sum_{\vec\theta} M(\vec\theta) \theta^2_y 
    & - \sum_{\vec\theta} M(\vec\theta) \theta_x \theta_y \\
    - \sum_{\vec\theta} M(\vec\theta) \theta_x \theta_y
    & \sum_{\vec\theta} M(\vec\theta) \theta^2_x
\end{matrix}
\right).
    \label{eqn:inertia}
\end{equation}
%
We always take $\lambda_1$ to be the largest value. As with $f_R$ and
$f_\theta$, we only consider mass up to one radial position passed the
outermost image and compute the fractional error as:
%
\begin{equation} \label{ferror shape}
  f_\mathrm{shape} = \left|s - \tilde s\right|  / \tilde s
\end{equation}
%
where $\tilde s$ is the shape of the mock galaxy. The distribution of
$f_\mathrm{shape}$ for each mock galaxy and each test case is shown in
\figref{shape results}.  Interestingly, for this global shape parameter
recovery it appears more important to have time delay data and/or a central
image (T,M) than to have a quad or multiple sources with wide redshift
separation. In all cases, the stellar mass little-aids the recovery, reflecting
the fact that $s$ is heavily weighted towards the shape at the {\it edge} of
the mass map, rather than at the centre where the stars may dominate the
potential (see \eqnref{eqn:inertia}).

>>>>

\title[Lens Recovery with \Glass]{Gravitational Lens Recovery with \Glass: Measuring the mass profile and shape of a lens}
\author[J.~P.~Coles~et~al.]{%
Jonathan P.~Coles,$^{1,3}$\thanks{jonathan@exascale-computing.eu}%
\newauthor%
Justin I.~Read$^2$,%
\newauthor%
and Prasenjit Saha$^3$%
%
\\
$^1$Exascale Research Computing Lab, Universit\'e de Versailles Saint-Quentin-en-Yvelines,\\$\quad$45 avenue des Etats-Unis, 78035, Versailles, France\\
$^2$Department of Physics, University of Surrey, Guildford, Surrey, GU2 7XH, United Kingdom \\
$^3$Physik-Institut, Universit\"at Z\"urich, 190 Winterthurerstrasse, 8057, Z\"urich, Switzerland
}

@@

To see why lens modelling details are of crucial importance, let us
recall the essential quantities that appear in lensing (see also \S\ref{sec:theory} for a more detailed exposition). First we have the distances. Let $D_L$, $D_S$, $D_{LS}$ be the
angular-diameter distances to the lens, source, and from lens to
source; these are all proportional to $c/H_0$ but have factors
that depend on the particular choice of cosmology\footnote{Here, $c$ is the speed of light in vacuo and $H_0$ is the Hubble parameter.}.  Typically:
%
\begin{equation}
D_L \approx z_L \frac c{H_0}\ \hbox{and}\ \frac{D_S}{D_{LS}} \sim 1.
\end{equation}
where $z_L$ is the redshift of the lens. 
%
For multiple images, the sky-projected density must exceed the critical
density in some region:
%
\begin{equation}
\Sigma_\mathrm{crit} = \frac{c^2}{4\pi G D_L} \sim \frac{1\rm\,kg\,m^{-2}}{z_L}
\end{equation}
where $G$ is Newton's gravitational constant. 
%
The angular separation between the lensed images is of order
the Einstein radius $\theta_E$, which is related to
the mass by:
%
\begin{equation}
\theta_E \sim \frac{R_G}{D_L} \frac{D_{LS}}{D_S}
\end{equation}
%
where $R_G = GM/c^2$ (with $M$ the projected mass enclosed within $\theta_E$)
is the gravitational radius. If the source is a quasar or otherwise rapidly
variable, a time delay $\Delta t$ in the variability will be present where
%
\begin{equation}
\Delta t \sim R_G/c
\end{equation}
%
So in principle, one can not only measure the mass of the lens, one can use the
dependence on the cosmology-dependent $D$ factors to extract the cosmological
model and all its parameters.  \cite{1937ApJ....86..217Z} drew attention to the
former, and \cite{1964MNRAS.128..307R,1966MNRAS.132..101R} pointed out the
latter, all long before lenses were discovered.  The difficulty with actually
doing this, however, became apparent soon after the discovery of the first lens
by \cite{1979Natur.279..381W}.  In the first ever paper on lens modelling,
\cite{1981ApJ...244..736Y} found that many plausible mass distributions could
reproduce the data.

@@

\subsection{A post-processing module for stellar kinematics}\label{sec:glasskinematics} 
Similarly to the null-space constraint (\secref{sec:glassextraimages}), stellar
kinematic constraints constitute a non-linear prior on the mass map and must be
applied in post-processing. We sweep through the model ensemble performing an
Abel deprojection to determine $\Mddd(r)$ from the projected surface density
$\Sigma(R)$ assuming spherical symmetry
\citep[e.g.][]{2008gady.book.....B,2008MNRAS.390.1647B}: 
%
\begin{eqnarray} 
    \Mddd(r) & = & M_\mathrm{p}(<r) - 4r^2 \int_0^{\pi/2} \Sigma\left(x\right) \left[\frac{1}{\cth[2]} \right. \nonumber \\ 
    & & \left. - \frac{\sth}{\cth[3]} \arctan\left(\frac{\cth}{\sth}\right) \right] \dth
\end{eqnarray}
%
where 
%
\begin{equation}
    M_\mathrm{p}(<r) = 2\pi \int_0^r R \Sigma(R) dR
\end{equation}
%
is the projected enclosed mass evaluated at 3D radius $r$; and $x = r/\cth$.

@@

\begin{figure*}
\includegraphics[width=0.33\textwidth]{BCQuadR1a_TmS-kappa-a.pdf}
\includegraphics[width=0.33\textwidth]{BCQuadR1a_TmS-kappa-b.pdf}
\caption{ \textbf{Left:} The mock data distribution for \mockBC{} projected
onto a coarse grid.  \textbf{Right:} The recovered ensemble average $\kappa$
distribution for the single quad with time delays. The contours are logarithmic
base 10 values, where level 0 corresponds to the critical density. Contours
below the critical density are drawn with dashed lines.}
\label{2d mass reconstruction}
\end{figure*}

@@

We can also more directly probe the recovery of the shape of the mass
distribution by defining a global measure of lens shape:
%
\begin{equation} 
    s \equiv \lambda_1/\lambda_2
\end{equation} 
%
where $\lambda_1$ and $\lambda_2$ are the eigenvalues of the 2D inertia tensor:
%
\begin{equation}
\left(
\begin{matrix}
   \sum_{\vec\theta} M(\vec\theta) \theta^2_y 
    & - \sum_{\vec\theta} M(\vec\theta) \theta_x \theta_y \\
    - \sum_{\vec\theta} M(\vec\theta) \theta_x \theta_y
    & \sum_{\vec\theta} M(\vec\theta) \theta^2_x
\end{matrix}
\right)
    \label{eqn:inertia}
\end{equation}
%
We always take $\lambda_1$ to be the largest value. As with $f_R$ and
$f_\theta$, we only consider mass up to one radial position passed the
outermost image and compute the fractional error as:
%
\begin{equation} \label{ferror shape}
  f_\mathrm{shape} = \left|s - \tilde s\right|  / \tilde s
\end{equation}
%
where $\tilde s$ is the shape of the mock galaxy. The distribution of
$f_\mathrm{shape}$ for each mock galaxy and each test case is shown in
\figref{shape results}.  Interestingly, for this global shape parameter
recovery it appears more important to have time delay data and/or a central
image (T,M) than to have a quad or multiple sources with wide redshift
separation. In all cases, the stellar mass little-aids the recovery, reflecting
the fact that $s$ is heavily weighted towards the shape at the {\it edge} of
the mass map, rather than at the centre where the stars may dominate the
potential (see \eqnref{eqn:inertia}).
